pdp: An R Package for Constructing Partial Dependence Plots

### 2016-12-28 manuscript/greenwell.tex
fixed bug in getTrainingData.train

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Introduction
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

@@

Constructing a PDP \eqref{eqn:pdf} in practice is rather straightforward. To simplify, let $\boldsymbol{z}_s = x_1$ be the predictor variable of interest with unique values $\left\{x_{11}, x_{12}, \dots, x_{1k}\right\}$. The partial dependence of the response on $x_1$ can be constructed as follows:
\begin{enumerate}
  \item For $i \in \left\{1, 2, \dots, k\right\}$:
  \begin{enumerate}
    \item Copy the training data and replace the original values of $x_1$ with the constant $x_{1i}$.
    \item Compute the vector of predicted values from the modified copy of the training data.
    \item Compute the average prediction to obtain $\bar{f}_1\left(x_{1i}\right)$.
  \end{enumerate}
  \item Plot the pairs $\left\{x_{1i}, \bar{f}_1\left(x_{1i}\right)\right\}$ for $i = 1, 2, \dotsc, k$.
\end{enumerate}
This can be quite computationally intensive since the algorithm involves $k$ passes over the training records. Fortunately, this can be parallelized quite easily (more on this in a later section). This algorithm can also be easily extended to larger subsets of two or more features as well.

@@

\subsection{Classification problems}

@@

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Introduction
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:intro}

@@

Constructing a PDP \eqref{eqn:pdf} in practice is rather straightforward. To simplify, let $\boldsymbol{z}_s = x_1$ be the predictor variable of interest with unique values $\left\{x_{11}, x_{12}, \dots, x_{1k}\right\}$. The partial dependence of the response on $x_1$ can be constructed as follows:

\begin{algorithm}
\begin{enumerate}
  \item For $i \in \left\{1, 2, \dots, k\right\}$:
  \begin{enumerate}
    \item Copy the training data and replace the original values of $x_1$ with the constant $x_{1i}$.
    \item Compute the vector of predicted values from the modified copy of the training data.
    \item Compute the average prediction to obtain $\bar{f}_1\left(x_{1i}\right)$.
  \end{enumerate}
  \item Plot the pairs $\left\{x_{1i}, \bar{f}_1\left(x_{1i}\right)\right\}$ for $i = 1, 2, \dotsc, k$.
\end{enumerate}
\caption{A simple algorithm for consructing the partial dependence of the response on a single predictor $x_1$. \label{alg:pdp}}
\end{algorithm}
Algorithm~\ref{alg:pdp} can be quite computationally intensive since it involves $k$ passes over the training records. Fortunately, this can be parallelized quite easily (more on this in a later section). This algorithm can also be easily extended to larger subsets of two or more features as well.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Classification problems
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Classification problems}
\label{sec:class}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% User-defined prediction functions
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{User-defined prediction functions}
\label{sec:prediction}

PDPs are essentially just averaged predictions; see, for example, step 1. (c) in Algorithm~\ref{alg:pdp}. As pointed at by \citet{goldstein-peeking-2015}, since traditional PDPs are based on the sample mean, strong interactions can conceal the complexity of the modeled relationship between the response and predictors of interest. This was part of the motivation behind \citeauthor*{goldstein-peeking-2015}'s ICE curves.

With \code{partial} it is possible to replace the mean in step 1. (c) with, say, the median, or obtain PDPs for classification problems on the raw probability scale. It is even possible to obtain ICE curves. This flexibility is due to the new \code{pred.fun} argument in \code{partial} (starting with \pkg{pdp} version 0.4.0). This argument accepts an optional prediction function that requires two arguments: \code{object} and \code{newdata}. The supplied prediction function must return either (i) a single prediction---typically an averaged prediction---or  (ii) a vector of predictions. The two examples below illustrate.

Using the \code{pred.fun} argument it is possible to obtain PDPs for classification problems on the probability scale; we just need to write a function that extracts the predicted class probabilities of interest and then averages the result. The snippet of code below uses the fitted SVM from Section~\ref{sec:class} to extract the average predicted probability of belonging to the \code{Setosa} class.
\begin{example}
pred.prob <- function(object, newdata) {  # see ?predict.svm
  pred <- predict(object, newdata, probability = TRUE)
  prob.setosa <- attr(pred, which = "probabilities")[, "setosa"]
  mean(prob.setosa)
}
\end{example}
Next, we simply pass this function via the \code{pred.fun} argument in the call to \code{partial}. The result is displayed in Figure~\ref{fig:partial_iris_prob}.
\begin{example}
# Figure 8
partial(iris.svm, pred.var = "Petal.Width", pred.fun = pred.prob, plot = TRUE,
        rug = TRUE)
\end{example}

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{partial_iris_svm_prob}
  \caption{Partial dependence of \code{Species} on \code{Petal.Width} and \code{Petal.Length} plotted on the probability scale; in this case, the probabiltity of belonging to the \code{Setosa} species.}
  \label{fig:partial_iris_prob}
\end{figure}

\newpage

For regression problems, the default prediction function is essentially
\begin{example}
pred.fun <- function(object, newdata) {
  mean(predict(object, newdata), na.rm = TRUE)
}
\end{example}
This corresponds to step step 1. (c) in Algorithm~\ref{alg:pdp}. Suppose we would like ICE curves instead. To accomplish this we simply need to pass a prediction function that returns a vector of predictions, one for each observation in \code{newdata}. This is demonstarted for the Boston housing example in the code snippet below.
\begin{example}
pred.ice <- function(object, newdata) {  # all predictions
  predict(object, newdata)
}
lstat.ice <- partial(boston.randomForest, pred.var = "lstat",
                     pred.fun = pred.ice)  # Figure 9 (left)
\end{example}

Recall that \code{partial} constructs a data frame that also inherits from class \code{"partial"}. When the function supplied to \code{pred.fun} returns multiple predictions, \code{partial} instead constructs a data frame that also inherits from class \code{"partial.ice"}. The only difference is the addition of a new column called \code{"obs"} which indicates which observation a curve belongs to; in this example, there is one curve for each unique observation in \code{boston}. To further illustrate, the code below constructs ICE curves for specific quantiles of the predictions. The results, along with the traditional PDP for \code{lstat}, are displayed in Figure~\ref{fig:partial_boston_ice}.
\begin{example}
pred.ice.quan <- function(object, newdata) {  # quantiles of predictions
  quantile(predict(object, newdata), probs = 1:9/10)
}
lstat.ice.quan <- partial(boston.randomForest, pred.var = "lstat",
                          pred.fun = pred.ice.quan)  # Figure 9 (middle)
\end{example}

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{partial_boston_ice}
  \caption{ICE plots and PDP for the Boston housing data. \textit{Left}: ICE curves. \textit{Middle}: Specific quantiles of ICE curves. \textit{Right}: Standard PDP.}
  \label{fig:partial_boston_ice}
\end{figure}
